Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
Outside the world of SMT, model finding has been
studied many times.

Some finite model finders are based on a reduction to Boolean satisfiability
(SAT). They translate the input problem to propositional logic and pass it to a
SAT solver. The translation is parameterized by upper or exact finite bounds on
the cardinalities of the atomic types. This procedure was pioneered by McCune
in the earlier versions of Mace (also capitalized MACE) \cite{mccune-1994}.
Other SAT-based finders are Paradox \cite{claessen-sorensson-2003}, FMDarwin
\cite{xxxx}, and iProver \cite{xxxx} for first-order logic with equality;
\ct{FMDarwin is not SAT-based although it is Mace-like.
Calling iProver SAT-based is a stretch since it is based 
on a natively first-order calculus even it uses an auxiliary SAT solver heavily.
}
Alloy Analyzer and its back-end Kodkod \cite{torlak-jackson-2007} for
first-order relational logic; and Refute \cite{weber-2008} and Nitpick
\cite{blanchette-nipkow-2010} for higher-order logic.

An alternative to translating the problem to propositional logic is to perform
an exhaustive model search directly on the original problem. Given fixed
cardinalities, the search space is represented in memory as multidimensional
arrays. The procedure tries different values in the function and predicate
table entries, checking each time if the problem is satisfied and backtracking
if necessary. This approach was pioneered by FINDER \cite{slaney-1994} and
SEM \cite{zhang-zhang-1995} and serves as the basis of many more model finders,
notably the Alloy Analyzer's precursor \cite{jackson-1996} and the later
versions of~Mace \cite{mccune-prover9-mace4}.

@@

The Leon system~\cite{blanc2013overview} implements a procedure that can both
produces proofs and counterexamples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth, which is increased iteratively as necessary.
The approach described in this paper works in an analagous manner, 
where instead the SMT solver is invoked only once 
and where quantifier instantiation amounts to function unfolding.
It is worth investigating how existing approaches for algorithms based on function unfolding
can inform approaches for dedicated quantifier instantiation techniques for function definitions,
and vice versa.
%\jb{Andy: Can you add some more here? You're the Leon pro! E.g. why/when do we think
%that our approach might be better?}
%\ct{About Leon, the reviewers and readers may wonder why we do not do the same, that is, just progressively unfold the recursive definitions until a model is found.
%We should probably say something in this regard.
%}
% It is not a question of "should you instead unfold function definitions?" since that's essential what quantifier instantiation does.  
% The question is "is quantifier instantiation a good way of unfolding function definitions?" 
% or "how can we instrument quantifier instantiation to mimic/improve upon existing strategies for function unfolding? --ajr

>>>>

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
The finite model finding support in the instantiation-based iProver \cite{xxxx}
is also closery related, given the similarities with SMT.

%Model finding has been studied outside the world of SMT.

Some finite model finders are based on a reduction to a decidable logic,
typically propositional logic. They translate the input problem to the weaker
logic and pass it to a solver for that logic.
The translation is parameterized by upper or exact finite bounds on
the cardinalities of the atomic types. This procedure was pioneered by McCune
in the earlier versions of Mace (also capitalized MACE) \cite{mccune-1994}.
Other conceptually similar finders are Paradox \cite{claessen-sorensson-2003}
and FM-Darwin \cite{baumgartner-et-al-2009} for first-order logic with
equality; the Alloy Analyzer and its back-end Kodkod \cite{torlak-jackson-2007}
for first-order relational logic; and Refute \cite{weber-2008} and Nitpick
\cite{blanchette-nipkow-2010} for higher-order logic.

An alternative to translating the problem is to perform
an exhaustive model search directly on the original problem. Given fixed
cardinalities, the search space is represented in memory as multidimensional
arrays. The procedure tries different values in the function and predicate
table entries, checking each time if the problem is satisfied and backtracking
if necessary. This approach was pioneered by FINDER \cite{slaney-1994} and
SEM \cite{zhang-zhang-1995} and serves as the basis of many more model finders,
notably the Alloy Analyzer's precursor \cite{jackson-1996} and the later
versions of~Mace \cite{mccune-prover9-mace4}.

@@

The Leon system~\cite{blanc2013overview} implements a procedure that can both
produces proofs and counterexamples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth, which is increased iteratively as necessary.
The approach described in this paper works in an analagous manner, 
where instead the SMT solver is invoked only once 
and where quantifier instantiation amounts to function unfolding. It is worth
investigating how existing approaches for algorithms based on function
unfolding can inform approaches for dedicated quantifier instantiation
techniques for function definitions, and vice versa.
